---
title: Software used for analysis and documentation
---

## Details of R session

```{r}
#| eval: true
#| warning: false
#| message: false

# spatial data
library(sf)

#map viz
library(leaflet)
library(tmap)
library(tmaptools)
#library(rnaturalearth)
library(OpenStreetMap)

# plots / graphics
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggforce)
library(grid)
library(DiagrammeR)

#tidyverse
library(dplyr)
library(magrittr)
library(lubridate)
library(tidyr)
library(stringr)


#models / ML
library(gbm)
library(caret)
library(dismo)

# other utils
library(here)
library(units)
library(RColorBrewer)

# documentation
library(quarto)
library(respondR)


```

All statistical analysis were performed using `r R.version$version.string` [@CRAN]. For the suitability model we used package *caret* [@pkg_caret] and some functions of packages s *gbm* [@pkg_gbm] and *dismo* [@pkg_dismo].

```{r}
sessionInfo()

```

## Other software and templates

Apart from R we used these pieces of software and templates:

- [Quarto](https://quarto.org) 
  - An open-source scientific and technical publishing system
- RLE-pdf, available [on GitHub](https://github.com/red-list-ecosystem/RLEpdf) 
  - Quarto extension for a template to generate a PDF with (pretty) LaTeX styling for RLE assessments:
- [WebPlotDigitizer](https://automeris.io/WebPlotDigitizer/)
  - Web based tool to extract data from plots, images, and maps
- [Color Blind Vision Simulator](https://pilestone.com/pages/color-blindness-simulator-1#) from pilestone.com
- [Error Propagation Calculator](https://astro.subhashbose.com/tools/error-propagation-calculator) by Subhash Bose
